Submission and Formatting Instructions for \\ International Conference on Machine Learning (ICML 2019)

### 2019-04-24 paper/icml_ai4good/main.tex
intro milestone 1 - focussing on getting EMOD to work now

\section{Introduction}
Ending the epidemics of AIDS, tuberculosis, malaria and other infectious diseases by 2030 is a key target within the Good Health \& Well-Being section of the UN Sustainable Development Goals\cite{}.

In contrast to AIDS, Malaria can be routinely cured thanks to the disposability of a variety of medical interventions. 
Issues are further aggravated due to the increasing emergence of parasite resistance to multiple drugs, including artemisinin, or even insecticides. Hence, despite decades of substantial international efforts, initial hopes of regional elimination or even global eradication have not been met. So difficult is Malaria control that, even today, about a quarter of a million children under the age of 5 succumb to it every year in Africa alone.

To reach the WHO’s target of reducing malaria incidence and mortality rates by at least 90\% by 2030, policymakers increasingly turn to evidence-based methods. However, the collection of large quantities of field data is often infeasible due to logistic, economic or political constraints, or outright impossible if novel methods are to be assessed a priori. Hence in silico simulation of malaria epidemiology plays an increasingly crucial role in evidence-based policymaking.
%Hence epidemiology simulators fine-tuned on real data play a crucial role in determining policy implications.

Even in the absence of political, economic and social constraints, Malaria epidemiology is governed by a complex set of drivers, few of which can be understood in isolation. These include within-host dynamics, population-specific traits and even local geography. Some progress in modeling specific components has been made, such as the parasite life-cycle, transmission vector dynamics and the inter-dependencies between infection rates and vicinity to open water. However, comprehensive modeling of all of these components in a regional context remains challenging, and is greatly impeded by the lack of available field data for regional calibration. Simulational complexity is further aggravated by considering feedback effects from a large arsenal of anti-malaria interventions, ranging from distribution of mosquito nets, indoor residual spraying and mass drug administration to age-specific vaccinations.

@@

Stochastic epidemiology simulators have already proven to be an invaluable tool to policymakers with policy implications that are often counterintuitive yet far-ranging. For example, it has been shown that mass vaccination may be largely ineffective in regions of large transmission rates, but may play a crucial role in areas of low transmittivity.

@@

Recent work by \citep{Baydin} introduces a fundamental paradigm change that allows to tackle the inverse problem directly. Using a novel software tool named \textit{pyprob}, \citep{Baydin} extract program execution traces from state-of-the-art particle decay simulation engine SHERPA\cite{}. Using latest techniques from universal probabilistic programming inference, the sampled execution traces are then used in order to characterise the posterior distribution of the simulator output. From these, regions in simulator input space can be associated with specified outcomes.

SHERPA, however, is different from leading epidemiology simulators in that it does not simulate population dynamics, but instead models individual decays at a time. In this paper, we for the first time demonstrate the applicability of \textit{pyprob}, and hence probabilistic programming inference, to population-based epidemiology simulators. In particular, we discuss how the leading malaria epidemiology simulators EMOD\cite{} and OpenMalaria\cite{} can be interfaced with \textit{pyprob}. We argue that our work is likely to have significant social impact on many livelihoods affected by Malaria and discuss opportunities for future research.

@@

\section*{References}

\bibliographystyle{icml2019}
\bibliography{refs}

>>>>

\section{Introduction}
Ending the epidemics of AIDS, tuberculosis, malaria and other infectious diseases by 2030 is a key target within the Good Health \& Well-Being section of the UN Sustainable Development Goals \cite{refugees_2030_nodate}\cite{un_sustainable_2018}.

In contrast to AIDS, Malaria can be routinely cured thanks to the disposability of a variety of medical interventions. However, both strands of parasite and associated mosquito host are ever-evolving and transmission vectors are hard to control. These circumstances are further aggravated due to the increasing emergence of parasite resistance to multiple drugs, including artemisinin, and even mosquito resistance to insecticides. In addition, many of the countries most affected by Malaria are less developed countries, in which logistic and issues of socio-economic stability often impede effective access to treatment.  Hence, despite decades of substantial international efforts, initial hopes of regional elimination or even global eradication have largely not been met. So difficult is Malaria control that, even today, about a quarter of a million children under the age of 5 succumb to it every year in Africa alone\cite{un_sustainable_2018}.

To reach the WHO’s target of reducing malaria incidence and mortality rates by at least 90\% by 2030, policymakers increasingly turn to evidence-based methods\cite{who_global_2015}. However, the collection of large quantities of field data is often infeasible due to logistic, economic or political constraints, or outright impossible if novel methods are to be assessed a priori. Hence in silico simulation of malaria epidemiology plays an increasingly crucial role in evidence-based policymaking \cite{mwendera_challenges_2019}.
%Hence epidemiology simulators fine-tuned on real data play a crucial role in determining policy implications.

Even in the absence of political, economic and social constraints, Malaria epidemiology is governed by a complex set of drivers, few of which can be understood in isolation. These include within-host dynamics, population-specific traits and even local geography. Some progress in modeling specific components has been made, such as the parasite life-cycle, transmission vector dynamics and the inter-dependencies between infection rates and vicinity to open water. However, comprehensive modeling of all of these components in a regional context remains challenging, and is greatly impeded by the lack of available field data for regional calibration. Simulational complexity is further aggravated by considering feedback effects from a large arsenal of anti-malaria interventions, ranging from distribution of mosquito nets, indoor residual spraying and mass drug administration to age-specific vaccinations \cite{smith_ensemble_2012}.

@@

Stochastic epidemiology simulators have already proven to be an invaluable tool to policymakers with policy implications that are often counterintuitive yet far-ranging. For example, it has been shown that mass vaccination may be largely ineffective in regions of large transmission rates, but may play a crucial role in areas of low transmittivity \cite{smith_ensemble_2012}.

@@

Recent work by \citep{gunes_baydin_efficient_2018} introduces a fundamental paradigm change that allows to tackle the inverse problem directly. Using a novel software tool named \textit{pyprob}, \citep{gunes_baydin_efficient_2018} extract program execution traces from state-of-the-art particle decay simulation engine SHERPA \cite{gleisberg_event_2008}. Using latest techniques from universal probabilistic programming inference, the sampled execution traces are then used in order to characterise the posterior distribution of the simulator output. From these, regions in simulator input space can be associated with specified outcomes.

SHERPA, however, is different from leading epidemiology simulators in that it does not simulate population dynamics, but instead models individual decays at a time. In this paper, we for the first time demonstrate the applicability of \textit{pyprob}, and hence probabilistic programming inference, to population-based epidemiology simulators. In particular, we discuss how the leading open-source malaria epidemiology simulators EMOD \cite{bershteyn_implementation_2018} and OpenMalaria \cite{smith_towards_2008}\cite{chitnis_periodically-forced_2012} can be interfaced with \textit{pyprob}. We argue that our work is likely to have significant social impact on many livelihoods affected by Malaria and discuss opportunities for future research.

@@

%\section*{References}

\bibliographystyle{icml2019}
\bibliography{malaria,refs}
